Rheology of Cubic Blue Phases

### 2013-06-27 trunk/papers/bprheo/bprheo.tex
Update bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2028 40876441-3496-0410-a613-a2fca35b2ae2

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up into a cholesteric helix, lying either along the
vorticity or the flow gradient direction according to the imposed flow.
BPI leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- very slow flow leads to a transition into an amorphous network 
with an apparent yield stress. The flow-induced oscillation regime features a novel arrangement
of double helical disclinations which rotate due to the shear. Large flow rates can, 
just as in the BPII case, lead to the rupture of the network and the stabilisation
of less ordered cholesteric structures. Our results provide the first 
theoretical investigation of sheared blue phases in large systems, and
are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
its helical axis, small pressure differences drive flow mainly through
``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase.
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

@@

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho v_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta [ \partial_\alpha v_\beta +\; \partial_\beta v_\alpha].
\label{NSE}
\end{eqnarray}
This emerges from the Chapman-Enskog expansion
of the lattice Boltzmann (LB) equations 
that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}$ that formally appears
in this expansion is negligible under the slow flows considered here
for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
The thermodynamic stress tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
\label{Pi}
\end{eqnarray}
and is responsible for strong non-Newtonian flow effects.
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} reduces to the
scalar pressure as would appear in Eq.~\ref{NSE} for a Newtonian fluid.

We next define a 
dimensionless number that describes the deformation
of the director field under flow. This so-called Ericksen number
is given by 
\begin{equation}
Er=\frac{\eta v l}{K}
\end{equation}
with $\eta$ and $K$ defined previously, and 
$v$ and $l$ a typical velocity and length scale. 
In the present work $l=p_0/2$ was used as this is the approximate size of
the BP unit cell. Likewise, $v$ was taken to be the velocity difference
across one unit cell, i.e. $v=\dot{\gamma} p_0/2$.

\if{The spectrum $X_\omega$ of the shear stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
\begin{equation}
X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N),
\label{spectrum}
\end{equation}
where we subtract the average stress $\bar{\Pi}_{xy}$ from the signal.
$N$ was the number of points in the time series, typically one for every 100 LB time steps. }\fi

@@

Simple shear flow was imposed by means of the Lees-Edwards boundary
conditions with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
For clarity we classify various flow regimes, two in the case of BPII and three in the 
case of BPI. 
They are BPII-1 ($\gd \lesssim 3.125\e{-4}$, corresponding to ${\it Er} \lesssim 2.67$), 
BPII-2 ($4.687\lesssim \gd\lesssim 6.25\e{-4}$; $4\lesssim {\it Er} \lesssim 5.34$), ),
BPI-1 ($\gd \lesssim 1.95\e{-5}$; ${\it Er} \lesssim 0.33$), ), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$; $0.67\lesssim {\it Er} \lesssim 8$) and BPI-3 ($6.25\e{-4}\lesssim \gd\lesssim 7.813\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$).
Beyond these regimes, and prior to the ultimate breakdown into an isotropic phase, 
both BPI and BPII form common configurations regardless 
of the initial state. These are a cholesteric helix with helical axis 
oriented along the direction of the velocity gradient (CHG), 
and a flow-aligned nematic state (FAN).  
As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd$ (normalised to $\eta$) 
of BPII versus time (top) and strain (bottom). 
For clarity the individual curves have been shifted along the time and strain axis. 
The inset in the top panel 
shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles), 
break up of the network into a travelling helical wave (BPII-2, solid circle), 
cholesteric helix with axis in gradient direction (CHG, open squares)
and flow-aligned nematic state (FAN, solid square).
The error bars indicate the maximum and minimum stresses that occur during one cycle.}
\label{bp2-rheo}
\end{figure}

We start our discussion with BPII as its flow behaviour is somewhat simpler 
than that of BPI. BPII has simple cubic symmetry and the disclination lines 
intersect and form a characteristic network of nodes.
Ahead of the more detailed discussion we provide first a general overview of the 
flow behaviour at all applied shear rates.
Fig. \ref{bp2-rheo} shows the ratio between the apparent viscosity, 
defined as 
$\eta_{app}=\langle \Pi_{xy} \rangle/\gd$, and $\eta$.
A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
without any additional contribution from the liquid crystal.
The top picture shows the data over time, whereas the graph at the bottom gives 
the same data versus accumulated strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is because of the periodic breakup and reconnecting of the network in shear flow. 
We refer to this regime as BPII-1. As noted in the Appendix, the
absence at very low shear rate of a permeation mode is due to our
choice of boundary conditions.
%{We note that permeation flows, which would
%{\em not} lead to oscillatory response, are effectively disallowed by the combination
%of the geometry and boundary condition we consider. This is
%because the periodic boundary conditions impose macroscopic distortion
%of the network, and are effective analogous to infinitely distant
%but anchored wall along the $xy$ plane.
%This means that  our results are effectively valid for sheared bulk 
%BPs which are anchored at the walls. Furthermore, permeation flows are 
%only restricted to very slow flows, and are unstable for intermediate 
%and high flows, where the response of the network depends less on
%the details of the boundary conditions used at 
%the wall~\cite{Marenduzzo:2006a,Marenduzzo:2006b}.}

The inset to Fig. 1 (top) shows a flow curve, defined as time averages of 
the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
$a=0.35, b=0.95$ describes the data to a very good approximation. 
Hence, the degree of shear-thinning is remarkably small in BPII, in
the range of shear rates which we have explored here.

For shear rates $4.68\e{-4}\lesssim\gd\lesssim6.25\e{-4}$ ($4\lesssim {\it Er} \lesssim 5.34$) the network of disclination lines
breaks up completely, and oscillations in the stress signal are absent.
In this regime of shear rates, to which we refer as BPII-2, the 
blue phase dissolves into a simple cholesteric liquid 
crystal. The helical axis is along the vorticity direction and
the flow excites travelling helical waves, such as those 
predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
the vorticity axis.

At shear rates $7.81\e{-4}\lesssim\gd \lesssim 9.37\e{-4}$ ($6.66\lesssim {\it Er} \lesssim 8$) the helix changes orientation.
The helical axis is now oriented along the gradient direction with
the liquid crystal flowing in ``nematic planes''. For even higher shear rates
the system undergoes a transition to a flow-aligned nematic state.

In the next sections we investigate the BPII-1 and BPII-2 flow regimes in more detail 
by looking at the kinetics of the disclination network.

@@

Fig. \ref{bp2-med} shows the disclination network in shear flow as
it undergoes homogeneous shearing in the BPII-1 flow regime. 
The disclination lines break up and 
reconnect further downstream, forming a periodically recurring pattern. The
period needed for a pattern to break up and 
reform along the flow direction is $\tau_F = 1/\gd$.

@@

Interestingly, while being displaced with the flow the entire network moves 
along the vorticity direction (z) as well.  A similar behaviour has been recently 
observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
In contrast to the stick-slip motion that has been reported there,
in this case the movement is steady. 
The vorticity motion occurs in such a way that the positions of breakup and reconnection 
point in the network, visible in Fig. \ref{bp2-med}, are slightly offset and allow the network 
to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
is $\tau_V=6\tau_F$, 
i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
for the network to move one unit cell along $z$ (the vorticity direction).

Fig. \ref{bp2-velo} shows the disclination network in the middle of a breakup-reconnection
cycle, with superimposed velocity vectors.
The plot shows the secondary velocity components, obtained by projecting the
velocity onto a plane perpendicular to the flow direction.
This gets rid of the dominating velocity component along the flow, $v_x$, and 
allows to visualise the patterns in the 
two much smaller components $v_y$ and $v_z$.
The magnitude of the secondary components is typically in the range of a 
few percent of the primary flow component, for the shear rates 
and system sizes simulated here.

\begin{figure}[htpb]
\center
\includegraphics[width=0.35\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.35\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.35\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: 
The pictures show a typical sequence of snapshots in the steady state 
at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
gradient is oriented along the vertical direction (y), whereas the 
horizontal direction (x) is the flow direction. Lees-Edwards boundary 
conditions have been imposed in such a way that the network moves to the 
right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure}

\begin{figure*}[htpb]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$.
The view is along the x-direction. This means the velocity ${\mathbf v}$ 
has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with right-handed helicity ($q_0>0$ in Eq.~\ref{FE}) travels rightwards, whereas the one one the right,
with reversed helicity, moves leftward. 
The periodicity of the motion along the vorticity direction is six times longer than the time 
it takes the network to reconnect along the flow direction.}
\label{bp2-velo}
\end{figure*}

Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
oriented along the vorticity direction. The direction of the flow
in each of the bands depends on the helicity of the underlying cholesteric
liquid crystal (i.e. on the sign of $q_0$ in Eq.~\ref{FE}), and the velocity
patterns for left-handed and right-handed BPs are one the mirror image of
the other. 
Further quantitative evidence for a direct link between the sense of motion 
and the helicity can be gained by time-averaging over individual cycles.

Table \ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. Therefore the averaged flow along the vorticity
direction, leading to network motion, is very small, and much smaller than
the local velocity flow. This suggests that the
flow of the network along the vorticity 
direction is permeative (i.e. the fluid moves but the network as a whole 
moves much less).

\subsubsection{Regime BPII-2: high shear rates }

If the flow rate exceeds a certain critical value BPII does not undergo the 
homogeneous deformation described above and adopts another mode of flow. 
Fig. \ref{bp2-high} shows a sequence of cuts through the director 
field at high shear rate. The entire blue phase network has broken up 
into a simple cholesteric helix and flows as a travelling 
helical wave with the helical axis oriented along the vorticity direction.
The state is translationally invariant in the flow and gradient directions.
It is the preferred mode of flow of a standard cholesteric phase
in vorticity mode at low Ericksen numbers. This is followed by
the uncoiling transition to a nematic, flow-aligned state at even 
higher shear rates ~\cite{Rey:1996a, Rey:1996b}.

In our scenario the BP to cholesteric transition occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
parameters for BPII ($K=0.02, \eta=0.666$, and $p_0/2=16$;
see Table \ref{tab1} in the Appendix).

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. The colour code gives the alignment of the director field in the flow direction (x) with red fully aligned and blue out of plane. 
The configuration is a helical wave that travels along the direction of vorticity.
It is translationally invariant along the gradient and flow direction.}
\label{bp2-high}
\end{figure}

We observe another configuration, lying between the travelling helical wave 
and the  flow-aligned nematic state at shear rates 
$7.81\e{-4}\lesssim\gd\lesssim9.37\e{-4}$ ($6.66\lesssim {\it Er} \lesssim 8$). In this interval
the helical axis now lies along the flow gradient direction (the so-called
Grandjean texture~\cite{deGennes}).
We will address this regime in section \ref{cholflow}, as it is common to BPI and BPII.

>>>>

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up 
\rev{
into a Grandjean texture with a 
cholesteric helix lying along the flow gradient direction according to the imposed flow.
}
BPI leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- very slow flow leads to a transition into an amorphous network 
with an apparent yield stress. Larger shear rates can make the network sensitive
to flow-induced transitions to metastable and amorphous states. For even
larger flow rates a Grandjean texture is observed, just as in the case of BPII.
At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.
Our results provide the first theoretical investigation of sheared blue phases in large systems, and are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
its helical axis, small pressure differences drive flow mainly through
``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
\rev{
Early experiments with cholesteric liquid crystals
showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
}
Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. 
\rev{
These result, however, were derived under the 
assumption that the molecules rotate only in flow-gradient plane whilst
the orientation of the cholesteric helix remained unchanged.
}  
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

@@

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho v_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta [ \partial_\alpha v_\beta +\; \partial_\beta v_\alpha].
\label{NSE}
\end{eqnarray}
This emerges from the Chapman-Enskog expansion
of the lattice Boltzmann (LB) equations 
that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}$ that formally appears
in this expansion is negligible under the slow flows considered here
for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.6666$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
The thermodynamic stress tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
\label{Pi}
\end{eqnarray}
and is responsible for strong non-Newtonian flow effects.
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} reduces to the
scalar pressure as would appear in Eq.~\ref{NSE} for a Newtonian fluid.

We next define a 
dimensionless number that describes the deformation
of the director field under flow. This so-called Ericksen number
is given by 
\begin{equation}
Er=\frac{\eta v l}{K}
\end{equation}
with $\eta$ and $K$ defined previously, and 
$v$ and $l$ a typical velocity and length scale. 
In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
the BP unit cell. Likewise, $v$ was taken to be the velocity difference
across one unit cell, i.e. $v=\dot{\gamma} l$. 
\rev{
The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.
}

@@

Simple shear flow was imposed by means of the Lees-Edwards boundary
conditions with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
\rev{
For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) that
may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
GJ and FAN occur regardless of the initial state.
}

\rev{
These regimes are characterised by the following flow rates: 
For Blue Phase II BPII-1 (PRC) ($\gd \lesssim 3.91\e{-4}$, corresponding to ${\it Er} \lesssim 3.33$), 
BPII-2 (GJ) ($4.69\lesssim \gd\lesssim 1.25\e{-3}; 4\lesssim {\it Er} \lesssim 10.67$), 
BPII-3 (FAN) ($1.875\e{-3}\lesssim \gd; 16 \lesssim {\it Er}$) 
and for Blue Phase I BPI-1 (AN) ($\gd \lesssim 1.95\e{-5}; {\it Er} \lesssim 0.17$),  
BPI-2 (PRC) ($3.91\e{-5}\lesssim \gd \lesssim 2.344\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$),
BPI-3 (AN) ($3.125\e{-4}\lesssim \gd\lesssim 4.69\e{-4}; 2.67 \lesssim {\it Er} \lesssim 4$),
BPI-4 (GJ and FGJ) ($5.47\e{-4} \lesssim \gd \lesssim 1.875\e{-3}; 4.67\lesssim {\it Er} \lesssim 16)$ and
BPI-5 (FAN) ($2.5\e{-3}\lesssim \gd, 21.33\lesssim {\it Er}$).
}

As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
\caption{
\rev{
Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
}
}
\label{bp2-flowcurve}
\end{figure}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
\caption{
\rev{
Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
of BPII versus strain $\gamma = \gd\cdot t$  
normalised to the background viscosity $\eta$.
The pictures show the regimes BPII-1 (top, PRC), BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.
}
} 
\label{bp2-appvisc}
\end{figure}

We start our discussion with BPII as its flow behaviour is somewhat simpler 
than that of BPI. BPII has simple cubic symmetry and the disclination lines 
intersect and form a characteristic network of nodes.
Ahead of the more detailed discussion we provide first a general overview of the 
flow behaviour at all applied shear rates.

Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
\rev{For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
the data to a very good approximation.
This holds true even in regime BPII-2, where the network breaks up and 
the liquid crystal adopts a Grandjean structure. The coefficents of the power law fit are
here $a=0.68$ and $b=0.94$. Hence, the degree of shear-thinning is 
remarkably small in BPII, in the range of shear rates which we have explored here.}

\rev{
Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
over total strain.
A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
without any additional contribution from the liquid crystal.
The top picture shows data for regime BPII-1, where the periodic breakup and reconnecting of the network 
in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
absence at very low shear rate of a permeation mode is probably due to our
choice of boundary conditions.
}

%{We note that permeation flows, which would
%{\em not} lead to oscillatory response, are effectively disallowed by the combination
%of the geometry and boundary condition we consider. This is
%because the periodic boundary conditions impose macroscopic distortion
%of the network, and are effective analogous to infinitely distant
%but anchored wall along the $xy$ plane.
%This means that  our results are effectively valid for sheared bulk 
%BPs which are anchored at the walls. Furthermore, permeation flows are 
%only restricted to very slow flows, and are unstable for intermediate 
%and high flows, where the response of the network depends less on
%the details of the boundary conditions used at 
%the wall~\cite{Marenduzzo:2006a,Marenduzzo:2006b}.}

\rev{
We observe another configuration, lying between those with periodically recurring conformations 
and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.
}

\rev{
The picture at the bottom of Fig.~\ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
the flow-aligned nematic state. We will address these regimes in section \ref{gj-fan}, as they are common to BPI and BPII.
}

\rev{
The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
the vorticity axis turns out to be actually a metastable state and features larger stress and higher dissipation due 
to the tumbling motion of the local director field, which is absent in the Grandjean texture.
For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
}

In the next sections we investigate the BPII-1 flow regime in more detail 
by looking at the kinetics of the disclination network.

@@

Fig. \ref{bp2-1-disc} shows the disclination network in shear flow as
it undergoes homogeneous shearing in the BPII-1 flow regime. 
The disclination lines break up and 
reconnect further downstream, forming a periodically recurring pattern. The
period needed for a pattern to break up and 
reform along the flow direction is $\tau_F = 1/\gd$.

@@

Interestingly, while being displaced with the flow the entire network moves 
along the vorticity direction (z) as well.  A similar behaviour has been recently 
observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
In contrast to the stick-slip motion that has been reported there,
in this case the movement is steady. 
The vorticity motion occurs in such a way that the positions of breakup and reconnection 
point in the network, visible in Fig. \ref{bp2-1-disc}, are slightly offset and allow the network 
to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
is $\tau_V=6\tau_F$, 
i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
for the network to move one unit cell along $z$ (the vorticity direction).

Fig. \ref{bp2-1-velo} shows the disclination network in the middle of a breakup-reconnection
cycle, with superimposed velocity vectors.
The plot shows the secondary velocity components, obtained by projecting the
velocity onto a plane perpendicular to the flow direction.
This gets rid of the dominating velocity component along the flow, $v_x$, and 
allows to visualise the patterns in the 
two much smaller components $v_y$ and $v_z$.
The magnitude of the secondary components is typically in the range of a 
few percent of the primary flow component, for the shear rates 
and system sizes simulated here.

\begin{figure}[htpb]
\center
\includegraphics[width=0.35\textwidth]{disc-160k_run902r.png}
\includegraphics[width=0.35\textwidth]{disc-164k_run902r.png}
\includegraphics[width=0.35\textwidth]{disc-165k_run902r.png}
\caption{Disclination network of BPII in shear flow: 
The pictures show a typical sequence of snapshots in the steady state 
at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
gradient is oriented along the vertical direction (y), whereas the 
horizontal direction (x) is the flow direction. Lees-Edwards boundary 
conditions have been imposed in such a way that the network moves to the 
right in the upper half and to the left in the lower half of the picture.}
\label{bp2-1-disc}
\end{figure}

\begin{figure*}[htpb]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902r.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903r.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$.
The view is along the x-direction. This means the velocity ${\mathbf v}$ 
has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with right-handed helicity ($q_0>0$ in Eq.~\ref{FE}) travels rightwards, whereas the one one the right,
with reversed helicity, moves leftward. 
The periodicity of the motion along the vorticity direction is six times longer than the time 
it takes the network to reconnect along the flow direction.}
\label{bp2-1-velo}
\end{figure*}

Characteristic bands are visible in Fig. \ref{bp2-1-velo}, which are 
oriented along the vorticity direction. The direction of the flow
in each of the bands depends on the helicity of the underlying cholesteric
liquid crystal (i.e. on the sign of $q_0$ in Eq.~\ref{FE}), and the velocity
patterns for left-handed and right-handed BPs are one the mirror image of
the other. 
Further quantitative evidence for a direct link between the sense of motion 
and the helicity can be gained by time-averaging over individual cycles.

Table~\ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. 
\rev{
Therefore the averaged flow along the vorticity
direction is very small, and much smaller than
the local velocity flow. This suggests that the
flow of the network along the vorticity 
direction is permeative. This claim is further supported by the characteristic 
velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
The flow in vorticity direction takes clearly place in positive {\em and} in negative z-direction, 
whereas the network as a whole moves only in one direction.
}
